---
title: "STAT 512 Homework 1"
author: "Kathleen Wendt"
date: "2/4/2020"
output: pdf_document
---

```{r packages, include = FALSE}
# load packages
library(tidyverse)
library(janitor)
library(GGally)
library(corrr)
library(kableExtra)
library(broom)
library(car)
```

```{r global_options, include = FALSE}
# set global options
knitr::opts_chunk$set(fig.width = 6, 
                      fig.height = 4, 
                      fig.path = "figs/",
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

Read Problem 12.53 (or 12.56 from the 6th Edition) which deals with cattle feed lot diets.

```{r cow_data}
# read cow data from 12.53
cow_data <- readr::read_csv("data/ex12-53.txt") %>% janitor::clean_names()
```

# Question 1: Scatterplots

Show the pairwise scatterplots between all 4 variables (Y=Time, X1=Protein, X2=Antibio, X3=Supplem).

```{r cow_pairs}
# 1. pairwise plot (base) for cow data
cow_data %>% 
  dplyr::select(-steer) %>% 
  plot()
```

\newpage

```{r cow_gg}
# 1. pairwise plot (gg) for cow data
GGally::ggpairs(cow_data, columns = c("time", "protein", "antibio", "supplem"))
```

# Question 2: Correlations

Calculate pairwise (Pearson) correlations between all 4 variables.

```{r cow_corr}
# 2. pearson correlations between variables of interest
cow_data %>% 
  dplyr::select(-steer) %>% 
  corrr::correlate() %>% 
  kableExtra::kable()
```

# Question 3: Simple linear regression model

Run the 3 simple linear regressions of Time vs each of the above three predictor variables. Show the parameter estimates (“Coefficients” table) and R2 values. You can just copy/paste the relevant output from R. (6 pts)

## Protein

```{r cow_lm_protein}
# 3a. simple lm - protein
lm_protein <- lm(time ~ protein, data = cow_data)
summary(lm_protein)
```

The $R^2$ value for the simple linear regression of time to market weight (Y) and protein (X) is 0.5057. See below for table of coefficient estimates and corresponding p-values.

```{r cow_lm_protein_tidy}
# 3a. create tidy lm df and table
lm_protein_tidy <- broom::tidy(lm_protein)
kableExtra::kable(lm_protein_tidy)
```

## Antibiotics

```{r cow_lm_antibio}
# 3b. simple lm - antibio
lm_antibio <- lm(time ~ antibio, data = cow_data)
summary(lm_antibio)
```

The $R^2$ value for the simple linear regression of time to market weight (Y) and antibiotics (X) is 0.1748. See below for table of coefficient estimates and corresponding p-values.

```{r cow_lm_antibio_tidy}
# 3b. create tidy lm df and table
lm_antiobio_tidy <- broom::tidy(lm_antibio)
kableExtra::kable(lm_antiobio_tidy)
```

## Supplements

```{r cow_lm_supplem}
# 3c. simple lm - supplem
lm_supplem <- lm(time ~ supplem, data = cow_data)
summary(lm_supplem)
```

The $R^2$ value for the simple linear regression of time to market weight (Y) and supplements (X) is 0.2203. See below for table of coefficient estimates and corresponding p-values.

```{r cow_lm_supplem_tidy}
# 3c. create tidy lm df and table
lm_supplem_tidy <- broom::tidy(lm_supplem)
kableExtra::kable(lm_supplem_tidy)
```

# Question 4: Multiple regression model

Now run multiple regression of Time on all three predictor variables. Show the parameter estimates (“Coefficients” table) and R2 value. We will use this the “full” model for the remaining questions.

```{r cow_multreg}
# 4. multiple regression with cow data
cow_multreg <- lm(time ~ protein + antibio + supplem, data = cow_data)
summary(cow_multreg)
```

The $R^2$ value for the multiple linear regression of protein, antibiotics, and supplements (X) on time to market weight (Y) is 0.9007. See below for table of coefficient estimates and corresponding p-values.

```{r cow_multreg_tidy}
# 4. create tidy lm df and table
cow_multreg_tidy <- broom::tidy(cow_multreg)
kableExtra::kable(cow_multreg_tidy)
```

# Question 5: Reflection (simple vs. multiple)

Note that (1) the slope estimates from the simple linear regressions are the same as the slope estimates from the “full” model and (2) the R2 values from the simple linear regressions sum to the R2 value from the “full” model. In general, this will not be the case (as we saw with the Rice Example). What is different about this data (as compared to the Rice Example)? Hint: Consider the result of question 2.

There is zero correlation between predictors (protein, antibiotics, supplements). Each predictor makes distinct contributions to the model. There is no multicollinearity among predictors.

# Question 6: Model assumptions and diagnostic plots

Create plots of (A) Residuals vs Fitted values and (B) QQplot of residuals. Include these plots in your assignment. Thinking about model assumptions, discuss your findings for each plot. (4 pts)

```{r cow_multreg_check}
# 6. visual check of regression assumptions
par(mfrow = c(2, 2))
plot(cow_multreg)
```

Assumptions:

- *Independence*: Unknown. Study design and experimentation are not described.

- *Linearity*: Assumed. The plot of residuals vs. fitted values does not show a trend (e.g., "megaphone" pattern) in residuals.

- *Equal variance*: Assumed. The plot of residuals vs. fitted values shows equal scatter among residuals.

- *Normality*: Assumed/acceptable. The Q-Q plot indicates an acceptable pattern of standardized residuals, considering the small sample size. 

# Question 7: Interpretation (R^2)

Interpret the R2 value from “full” model.

The $R^2$ value of the full multiple regression model indicates that 90.07% of the variation in time to market weight can be explained by the levels of protein, antibiotics, and supplements in cattle feed.

# Question 8: Interpretation (antibiotics)

Give a one-sentence interpretation of estimated partial regression coefficient for AntiBio in the multiple regression.

For every one-unit increase in antibiotic content in cattle feed, there is an estimated four-day decrease in time to market weight, holding all other variables (i.e., protein, supplements) constant.

# Question 9: Hypothesis tests (betas)

Working from the “full” model, for each of the four betas (intercept and three partial regression coefficients) give a p-value for the hypothesis that the true parameter value is zero vs a two-sided alternative. In other words, test null: beta-i is 0 versus alternative: beta-i is not 0.

```{r cow_multreg_tidy2}
# 9. create tidy lm df and table
cow_multreg_tidy <- broom::tidy(cow_multreg)
kableExtra::kable(cow_multreg_tidy)
```

Hypothesis tests of $\beta$ = 0 (null) vs. $\beta$ not 0 (alternative):

- Intercept: _p_ = `r cow_multreg_tidy$p.value[1]` < $\alpha$ = 0.05. Reject.
- Protein: _p_ = `r cow_multreg_tidy$p.value[2]` < $\alpha$ = 0.05. Reject.
- Antibiotics: _p_ = `r cow_multreg_tidy$p.value[3]` < $\alpha$ = 0.05. Reject.
- Supplements: _p_ = `r cow_multreg_tidy$p.value[1]` < $\alpha$ = 0.05. Reject.

# Question 10: Hypothesis test (protein)

Working from the “full” model, test the null hypothesis that the partial regression coefficient for Protein equals -3.0 versus a two-sided alternative. In other words, test $H_0$: $\beta$ = -3 vs. $H_A$: $\beta$ $\neq$ -3. Give a test statistic, p-value and conclusion. (4 pts) 

_Note_: One approach to this question uses the car package. Remember you need to install a package the first time you use it and load the package every time you use it!

```{r cow_protein_alt}
# 10. test protein alternative
protein_matrix <- c(0, 1, 0, 0)
protein_alt <- broom::tidy(car::lht(cow_multreg, protein_matrix, rhs = -3.0))
```

We reject the null hypothesis that the partial regression coefficient for protein is equal to -3. There is evidence to suggest it is not equal to -3, _F_ = `r protein_alt$statistic[2]`, _p_ = `r protein_alt$p.value[2]` < $\alpha$ = 0.05. 

\newpage

# Appendix

```{r show_code, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
