---
title: "Exam 1 Extra Practice"
author: "Kathleen Wendt"
date: "3/1/2020"
output: pdf_document
---

```{r packages, include = FALSE}
# load packages
library(tidyverse)
library(janitor)
library(car)
library(MuMIn)
```

```{r global_options, include = FALSE}
# set global options
knitr::opts_chunk$set(fig.width = 6, 
                      fig.height = 4, 
                      fig.path = "figs/",
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

# Question 1: Body Fat

```{r fat_data}
fat_data <- readr::read_csv("data/BodyFat.csv") %>% janitor::clean_names()
```

## 1A: VIF

```{r fat_full_lm}
# fit full fat model
fat_full_lm <- lm(body_fat ~ triceps + thigh + midarm, data = fat_data)

# calculate vif values for each predictor
car::vif(fat_full_lm)

# calculate vif "by hand" for triceps vs others - which slide??? from key:
1/(1-0.9986)
```

I don't know how to calculate VIF by hand or where to find the information in the slides.

## 1B: Refit

```{r fat_back_lm}
# fit fat model selected by backwards elimination
fat_back_lm <- lm(body_fat ~ triceps + midarm, data = fat_data)

# calculate vif values for each predictor
car::vif(fat_back_lm)
```

## 1C: Influence

```{r fat_cook}
# identify obs with largest cook's distance
plot(fat_back_lm)
influence.measures(fat_back_lm) # obs 3 largest

# consider if obs 3 is influential 
## dfbetas - abs value exceeds 2/sq(n)
2/sqrt(20)

## dffits - if abs value exceeds 2*sqrt((k+1)/n)
2*sqrt((2+1)/20)

## cooksd - if > 1
```

Observation 3 is influential based on DFBETAS and DFFITS but not according to Cook's distance.

# Question 2: Mortality

```{r death_data}
death_data <- readr::read_csv("data/Mortality.csv") %>%
  janitor::clean_names() %>% 
  dplyr::mutate(
    hc_log = log(hc),
    nox_log = log(nox),
    so2_log = log(so2)
  )
```

## 2A: Correlations

```{r death_corr}
death_data %>% 
  dplyr::select(mortality, hc_log, nox_log, so2_log) %>% 
  corrr::correlate()
```
There is a high correlation between hydrocarbon and nitrogen oxide pollution.

## 2B: Model selection

```{r death_lm}
# build mr with demo and climate predictors
death_demo_lm <- lm(mortality ~ 
                 precip + humidity + 
                 jantemp + julytemp + 
                 over65 + house + educ + 
                 sound + density + nonwhite + 
                 whitecol + poor, 
               data = death_data)

# select best model 
options(na.action = "na.fail")
death_demo_subset <- MuMIn::dredge(death_demo_lm, rank = "AIC")
```

The model selected includes `density`, `educ`, `house`, `jantemp`, `julytemp`, `nonwhite`, and `precip`.

## 2C: Add pollution predictors

```{r death_full_lm}
# update multreg with pollution indicators
death_full_lm <- lm(mortality ~
                      precip + jantemp + julytemp + house +
                      educ + density + nonwhite + 
                      hc_log + nox_log, so2_log,
                    data = death_data)

# select best model 
options(na.action = "na.fail")
death_poll_subset <- MuMIn::dredge(death_full_lm,
                                   fixed = c("density", "educ", "house",
                                             "jantemp", "julytemp", 
                                             "nonwhite", "precip"),
                                   rank = "AIC")

death_full_lm <- lm(mortality ~
                      precip + jantemp + julytemp + house +
                      educ + density + nonwhite + 
                      hc_log + nox_log,
                    data = death_data)
```

## 2D: VIF

```{r death_vif}
car::vif(death_full_lm)
```

There is collinearity between `hc_log` and `nox_log`.

## 2E: Interpretation

```{r death_summary}
summary(death_full_lm)
```

A one-unit increase in `hc_log` is associated with a 28.48 unit _decrease_ in predicted mortality, holding other variables in the model constant.

A one-unit increase in `nox_log` is associated with a 37.65 unit _increase_ in predicted mortality, holding other variables in the model constant.

