---
title: "STAT 512 Final Project"
subtitle: "Effects of Brief Meditation on Mother Respiratory Sinus Arrhythmia"
author: "Kathleen E. Wendt"
date: "May 10, 2020"
output: pdf_document
---

```{r global_options, include = FALSE}
# set global options for figures, code, warnings, and messages
knitr::opts_chunk$set(fig.width = 6, 
                      fig.height = 4, 
                      fig.path = "figs/",
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r load_packages, include = FALSE}
# load packages
library(tidyverse) # for reading and manipulating data
library(rstatix) # for pipe-friendly stat stuff
library(kableExtra) # for tables
library(nlme) # for repeated measures modeling
library(emmeans) # for comparing estimated marginal means
```

```{r prepare_data, include = FALSE}
# read in treatment assignment data and filter to mothers only
trt_audio <- readr::read_csv("data/audio_parent.csv") %>%
  dplyr::filter(individual == 1)
# filter to baseline rsa for each person
rsa_base <- readr::read_csv("data/hrv_parent_valid.csv") %>%
  dplyr::select(family, individual, task, rsa_mean_task) %>% 
  dplyr::filter(individual == 1,
                task == 1) %>% 
  dplyr::distinct() %>% 
  dplyr::select(-individual, -task) %>% 
  dplyr::rename(rsa_baseline = rsa_mean_task)
# prepare all data for summaries and models
rsa_trt_data <- readr::read_csv("data/hrv_parent_valid.csv") %>%
  dplyr::select(family, individual, task, rsa_mean_task) %>% 
  dplyr::filter(individual == 1) %>% 
  dplyr::distinct() %>% 
  dplyr::left_join(rsa_base, by = "family") %>%
  dplyr::full_join(trt_audio, by = c("family", "individual")) %>% 
  dplyr::select(-individual) %>% 
  dplyr::group_by(family) %>% 
  dplyr::mutate(rsa_diff = rsa_mean_task - rsa_baseline) %>% 
  dplyr::filter(task != 1) %>% 
  dplyr::ungroup() %>% 
  rstatix::convert_as_factor(family, task, condition) %>% 
  dplyr::rename(mother_id = family) ## rename family id as mother id
```

# Introduction

## Background

This project is a sliver of my master's thesis, which I will defend in June 2020. We conducted a pilot study of a brief mindfulness induction (focused-attention meditation) on self- and co-regulation between parents and young children (ages 4.5 - 6.5) during a challenging dyadic task. My master's thesis investigates the effects of a meditation on parent autonomic nervous system activity during a challenging parenting moment and individual differences in such by prior contemplative experience and dispositional mindfulness. 

## Research Question

General: How does a mindfulness meditation affect parasympathetic nervous system activity in mothers during a subsequent challenging parent-child interaction? 

Specific: How does the average change in respiratory sinus arrhythmia (response) in mothers (subject/block) differ by treatment group (meditation vs. podcast) across phases of a challenging parenting task (time period)?

# Description

The observational/experimental unit is an individual parent. 

For the purposes of this assignment, I only considered the subset of mothers (N = 40), but 27 fathers also participated in this study. The blocking variable was subject (unique mother identification as a factor).

The response variable was the average change in respiratory sinus arrhythmia (RSA), which is a noninvasive indicator of parasympathetic nervous system activity. RSA was a continuous variable but will be calculated as a difference score from baseline. 

The predictor variables were audio recording as treatment (between-subjects factor) and protocol task as time period (within-subjects factor). Treatment was a categorical variable with two levels: control (0) and experimental (1). Mothers were randomly assigned to a group, balancing for child gender, child age, parent gender, parent mental health, and family strengthening activities. The experimental group listened a focused-attention meditation, whereas the control group listened to an educational podcast about the benefits of sleep for the same duration. Time period was a categorical variable with 5 levels: 3-minute resting period; 5.5-minute audio treatment; 4-minute dyadic challenge (part 1); 3-minute dyadic stressor (part 2); 3-minute dyadic recovery (part 3). After the response is calculated as a difference score from baseline/resting period, only the latter 4 levels remained.

# Summary

```{r sum_stats, include = FALSE}
# rstatix sum stats
rsa_trt_data %>% 
  dplyr::select(-rsa_mean_task, -rsa_baseline) %>% 
  dplyr::group_by(task, condition) %>% 
  rstatix::get_summary_stats() %>% 
  dplyr::ungroup()
```

```{r boxplot}
# boxplot of mean rsa diff by time and trt
rsa_trt_data %>% 
  ggplot(aes(x = task, y = rsa_diff, color = condition)) + 
  geom_boxplot() +
  theme_bw() + 
  ggtitle("Mean RSA change across time by treatment")
```

```{r lineplot}
# line plot mean rsa diff score by time and trt
rsa_trt_data %>%
  ggplot(aes(x = task, y = rsa_diff)) + 
  stat_summary(aes(group = condition, color = condition), 
               geom = "line", fun = mean) + 
  theme_bw() + 
  ggtitle("Mean RSA change across time by treatment") 
```
# Pre-analysis assumptions

## Outliers

```{r outliers, include = FALSE}
rsa_trt_data %>% 
  dplyr::group_by(condition, task) %>% 
  rstatix::identify_outliers(rsa_diff)
```

There are no extreme outlying scores of RSA, based on inter-quartile calculations. 

## Normality

```{r normality, include = FALSE}
rsa_trt_data %>% 
  dplyr::group_by(condition, task) %>% 
  rstatix::shapiro_test(rsa_diff)
```

Based on the Shapiro-Wilk test, the residuals of the difference scores for RSA by time period (task) and treatment group (condition) are normally distributed.

# Analysis

Analysis was executed in `r R.version.string`, using the `nlme` and `emmeans` packages. A balanced two-factor (two-way) repeated measures analysis of variance (ANOVA) was performed to evaluate the effect of treatment across time on mean change in RSA. In order to account for intra-individual variability in RSA, I first calculated a difference score for each individual based on the baseline/resting average RSA. This difference score was then used as the response variable in the model. Fixed effects included treatment (experimental vs. control), time period (audio, 3-part parent-child task), and a treatment*time interaction. Subject (individual mother) was included as a random effect to account for repeated measures.

First, the assumption of compound symmetry was investigated. Three models were fit to be compared using AIC: basic, unstructured covariance, and autogressive covariance approaches. According to AIC, the model with autoregressive covariance fits the data most appropriately.

estimate/compare means corresponding to fixed effects

```{r model1, include = FALSE}
# MODEL 1: basic/corCompSymm: covariance same across period on same subject
options(contrasts = c("contr.sum", "contr.poly"))
model1_basic <- nlme::lme(rsa_diff ~ task*condition,
                                      random = ~1 | mother_id,
                                      data = rsa_trt_data)
summary(model1_basic)
nlme::anova.lme(model1_basic, type = "marginal")
emmeans::emmeans(model1_basic, dunnett ~ task|condition)
emmeans::emmeans(model1_basic, pairwise ~ task|condition)
```

```{r model2, include = FALSE}
# MODEL 2: corSymm: unstructured covariance; correlations allowed to differ
options(contrasts = c("contr.sum", "contr.poly"))
model2_corsymm <- nlme::lme(rsa_diff ~ task*condition,
                            random = ~1|mother_id,
                            correlation = corSymm(form = ~1|mother_id),
                            data = rsa_trt_data)
summary(model2_corsymm)
nlme::getVarCov(model2_corsymm, individual = 1, type = "marginal")
```

```{r model3, include = FALSE}
# MODEL 3: corAR1: autoregressive covariance; assumes covariance decreases 
# exp; assumes equal time spacing (violated with different task lengths?)
options(contrasts = c("contr.sum", "contr.poly"))
model3_corar1 <- nlme::lme(rsa_diff ~ task*condition,
                           random = ~1|mother_id,
                           correlation = corAR1(form = ~1|mother_id),
                           data = rsa_trt_data)
summary(model3_corar1)
nlme::anova.lme(model3_corar1, type = "marginal")
emmeans::emmeans(model3_corar1, dunnett ~ task|condition)
emmeans::emmeans(model3_corar1, pairwise ~ condition|task)
```

```{r compare}
AIC(model1_basic, model2_corsymm, model3_corar1)
```

## Results

- Type 3 ANOVA table and/or table of estimated coefficients or other tables

```{r}
# # Visualization: box plots with p-values
# pwc <- pwc %>% add_xy_position(x = "time")
# bxp +
#   stat_pvalue_manual(pwc, tip.length = 0, hide.ns = TRUE) +
#   labs(
#     subtitle = get_test_label(res.aov, detailed = TRUE),
#     caption = get_pwc_label(pwc)
#   )
```

# Conclusions 

- interpretation and discussion (even if nothing is significant!)
- address research questions 

There was a statistically significant interaction between treatment and time on self-esteem score, F(2, 22) = 30.4, p < 0.0001. Therefore, the effect of treatment variable was analyzed at each time point. P-values were adjusted using the Bonferroni multiple testing correction method. The effect of treatment was significant at t2 (p = 0.036) and t3 (p = 0.00051) but not at the time point t1 (p = 1).

Pairwise comparisons, using paired t-test, show that the mean self-esteem score was significantly different between ctr and Diet trial at time points t2 (p = 0.012) and t3 (p = 0.00017) but not at t1 (p = 0.55).

\newpage

# Appendix

```{r show_code, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
