---
title: "TITLE"
subtitle: "SUBTITLE"
author: "Kathleen Wendt"
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{indentfirst}
date: "May 10, 2020"
output: pdf_document
indent: true
---

```{r packages, include = FALSE}
# load packages
# library(readxl)
# library(tidyverse)
# library(janitor)
# library(car)
# library(emmeans)
```

```{r global_options, include = FALSE}
# set global options
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

The write-up should be 5 pages or fewer and written in the same style as a journal article, but with more detail and focus on the analysis. 

# Introduction

## Background

## Research question

1.

# Data description

## Observational/experimental units

## Sample size

## Response and predictors

- continuous or categorical (with levels)

## Study design

# Summary statistics and graphics

Typically done BEFORE formal model fitting

I do not expect the graphs to be “publication quality”, but they should be clearly labeled. As a rule, a graph should only be included in a paper if you have something to say about it.

# Analysis

approach from this class: MR, factorial ANOVA, mixed model

2+ predictors with single response variable (running same analysis for multiple response variables will not earn additional credit)

- description/discussion of analysis - reproducible

- justify any decisions (e.g., median-split)

- discuss model assumptions (consider including diagnostic plots)

# Results and conclusions

## Present results

- Type 3 ANOVA table and/or table of estimated coefficients or other tables
- interpretation and discussion (even if nothing is significant!)
- address research questions 

# STYLE 

- about 5 pages including graphs but excluding code or references
- writing quality 
- clean output (kable tables or in-line coding)
- code in appendix; congruency between code and description

# Appendix

```{r show_code, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
