---
title: "STAT 512 Homework 9"
author: "Kathleen Wendt"
date: "04/21/2020"
output: pdf_document
---

```{r med_packages, include = FALSE}
# load packages
library(readxl)
library(tidyverse)
library(janitor)
library(car)
library(emmeans)
```

```{r global_options, include = FALSE}
# set global options
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

# Part 1: Antihistamine data

For this problem use the data described in Ott & Longnecker problem 19.23 (p 1080 in the 7th Edition). We will verify that this is a BIBD in parts A and B. Note that Person acts as the blocking variable.

```{r allergy_data}
# read and prepare allergy/antihistamine data
allergy_data <- readxl::read_xlsx("data/ex19-23.xlsx") %>%
  janitor::clean_names() %>% 
  dplyr::mutate(person = as.factor(person),
                treatments = as.factor(treatments))
```

## Question 1: BIBD components

Identify t, r, b and k. (4 pts)

t = 6 antihistamine treatments 

r = 5 replicates of each treatment

b = 10 blocks (patients)

k = 3 natural block size (treatments per patient)

N = 30 observations

tr = bk 

30 = 30

## Question 2: BIBD lambda

Compute $\lambda$ (and make sure it is an integer).

$\lambda$ = r(k-1)/(t-1) = 5(3-1)/(6-1) = 10/5 = 2

Based on Q1 and Q2, this is a BIBD.

## Question 3: BIBD with block as fixed

Treat Person (block) as fixed. Fit an appropriate model (using `lm()`) and include the Type 3 ANOVA table (using `Anova(, type = 3)`) in your assignment.

```{r 3_med_fixed_bibd}
# 3. treat person as fixed in bibd lm 
med_bibd_fixed <- lm(area_red ~ treatments + person, data = allergy_data)
# 3. call anova type 3 table on fixed bibd lm
car::Anova(med_bibd_fixed, type = 3)
```

## Question 4: Fixed BIBD CLD

Calculate `emmeans` and Tukey adjusted p-values for their differences. Summarize using a `cld` display and include it in your assignment. (4 pts)

```{r 4_med_emmeans}
# 4. emmeans for fixed bibd lm
med_bibd_fixed_em <- emmeans::emmeans(med_bibd_fixed, pairwise ~ treatments)
# 4. cld for emmeans
emmeans::cld(med_bibd_fixed_em$emmeans)
```

## Question 5: Tukey's HSD

Calculate Tukey’s HSD value by hand. 

Tukey's HSD = 12.63

See code appendix for hand calculations.

```{r 5_med_hsd, include = FALSE}
# 5. calculate tukey critical value in R
qtukey(0.95, 6, 15)
# 5. calculate se diff in means "by hand"
sqrt(2*3*(453.61/15)/(6*2))
# 5. calculate tukey hsd "by hand"
4.595*(sqrt((453.61/15)/4))
# 5. short-cut with se from contrasts output
4.595*3.89/sqrt(2)
```

## Question 6: BIBD with block as random

```{r med_packages2, include = FALSE}
# detach emmeans
detach(package:emmeans)
# load random effects testing packages; lmer masked: lme4 -> lmerTest
library(lme4)
library(lmerTest)
# reload emmeans to work correctly with lmer
library(emmeans)
```

Treat Person (block) as random. Fit an appropriate model (using `lmer()` from `lme4` package) and include the Type 3 ANOVA table (using `anova(, ddf=”Kenward-Roger”)`).

```{r 6_med_bibd_rand}
# 6. treat person as random in allergy bibd 
med_bibd_rand <- lmer(area_red ~ treatments + (1|person), data = allergy_data)
# 6. call anova type 3 table on bibd random
anova(med_bibd_rand, ddf = "Kenward-Roger")
```

## Question 7: Random BIBD CLD

Calculate emmeans and Tukey adjusted p-values for their differences. Summarize using a cld display and include it in your assignment. (4 pts)

```{r 7_med_bibd_rand_em}
# 7. emmeans for random effects bibd lm
med_bibd_rand_em <- emmeans::emmeans(med_bibd_rand, pairwise ~ treatments)
# 7. cld for emmeans
emmeans::cld(med_bibd_rand_em$emmeans)
```

## Question 8: Fixed vs. random 

Comparing your results from #4 and #7, you should find that the emmeans are not the same. Is the ranking of the treatments the same? Are the conclusions about significant differences the same?

The estimated marginals means differ between the fixed effects and random effects models, but the ranking of the treatments and the conclusions about differences between treatments are the same.

# Part 2: Wheat variety data

In a variety trial, a total of 7 varieties (Var) of wheat are considered. For each variety, n = 3 reps are randomly assigned to positions in a single field. The response variable is the Yield at the end of the trial. The data is available from Canvas as “Varieties.csv”. 

```{r wheat_packages, include = FALSE}
# detach all additional packages in session thus far
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
# load packages (again)
library(tidyverse)
library(janitor)
library(car)
library(emmeans)
library(kableExtra)
```

```{r wheat_data}
# read and prepare wheat variety data
wheat_data <- readr::read_csv("data/Varieties.csv") %>% 
  janitor::clean_names() %>% 
  dplyr::mutate(var = as.factor(var))
```

## Question 9: Fixed effects model

Treat Variety as fixed and fit a one-way ANOVA model using `lm()`. Fit an appropriate model and construct the ANOVA table (using `Anova(, type = 3)`).

```{r 9_wheat_fixed_lm}
# 9. treat variety as fixed in wheat lm 
wheat_fixed_lm <- lm(yield ~ var, data = wheat_data)
# 9. call anova type 3 table on fixed wheat lm
car::Anova(wheat_fixed_lm, type = 3)
```

## Question 10: Null hypothesis

In the ANOVA table, you should find a test corresponding to Variety. Specifically, F = 18.26 and p-value <.0001. State the null hypothesis for this test.

No difference in yield by wheat variety.

## Question 11: Estimated marginal means

Give the emmeans for each variety.

```{r 11_wheat_fixed_lm_em}
# 11. compute emmeans
wheat_lm_em <- emmeans::emmeans(wheat_fixed_lm, pairwise ~ var)
# 11. display emmeans
kableExtra::kable(wheat_lm_em$emmeans)
```

## Question 12: Random effects model

Treat Variety as random and fit a one-way random effects model using `lmer()`. Fit an appropriate model and include “Random Effects” table (giving the variance component estimates) in your assignment.

```{r wheat_packages2, include = FALSE}
# detach emmeans
detach(package:emmeans)
# load random effects testing packages; lmer masked: lme4 -> lmerTest
library(lme4)
library(lmerTest)
# reload emmeans to work correctly with lmer
library(emmeans)
```

```{r 12_wheat_rand_lm}
# 12. treat variety as random in wheat lm 
wheat_rand_lm <- lmer(yield ~ (1|var), data = wheat_data)
# 12. call summary on random wheat lm; type 3 anova table empty because 
# intercept is only fixed effect
summary(wheat_rand_lm)
```

## Question 13: Variation from variety

Use the `rand()` function to get a test corresponding to Variety. Specifically, Chi.sq = 19 and p-value < 0.0001. State the null hypothesis for this test.

```{r 13_wheat_rand_var, include = FALSE}
# 13. test potency/value of random-effect term (variety) in lm
lmerTest::rand(wheat_rand_lm)
```

Null hypothesis: No additional variation from wheat type; $\sigma^2$ for variety = 0. 

## Question 14: Best Linear Unbiased Predictors

Give the BLUPs for each variety. Remember to include the intercept in the BLUP estimation.

```{r 14_wheat_rand_blups}
# 14. compute blups for wheat variety
lme4::ranef(wheat_rand_lm)
```

\newpage

# Appendix

```{r show_code, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
