---
title: "STAT 512 Homework 9"
author: "Kathleen Wendt"
date: "04/28/2020??"
output: pdf_document
---

```{r packages, include = FALSE}
# load packages
library(tidyverse)
library(janitor)
# library(kableExtra)
# library(car)
# library(broom)
# library(emmeans)
```

```{r global_options, include = FALSE}
# set global options
knitr::opts_chunk$set(fig.width = 6, 
                      fig.height = 4, 
                      fig.path = "figs/",
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

# Part 1: O & L data

For this problem use the data described in Ott & Longnecker problem 19.23 (p 1080 in the 7th Edition). We will verify that this is a BIBD in parts A, B. Note that Person acts as the blocking variable.

## Question 1

Identify t, r, b and k. (4 pts) Hint: Use the `table()` function.

## Question 2

Compute $\lambda$ (and make sure it is an integer).

## Question 3

Treat Person (block) as fixed. Fit an appropriate model (using `lm()`) and include the Type 3 ANOVA table (using `Anova(, type = 3)`) in your assignment.

## Question 4

Treat Person (block) as fixed. Calculate `emmeans` and Tukey adjusted p-values for their differences. Summarize using a `cld` display and include it in your assignment. (4 pts)

## Question 5

Treat Person (block) as fixed. Calculate Tukey’s HSD value by hand. Hint: See the ExpDesign 3 slide 104. Recall that you can check your calculation using the SE value from the previous question.

## Question 6

_Note: Be sure to detach the emmeans package. Then load the lme4, lmerTest, pbkrtest and emmeans packages._

Treat Person (block) as random. Fit an appropriate model (using `lmer()` from `lme4` package) and include the Type 3 ANOVA table (using `anova(, ddf=”Kenward-Roger”)`)

## Question 7

Treat Person (block) as random. Calculate emmeans and Tukey adjusted p-values for their differences. Summarize using a cld display and include it in your assignment. (4 pts)

## Question 8

Treat Person (block) as random. Comparing your results from #4 and #7, you should find that the emmeans are not the same. Is the ranking of the treatments the same? Are the conclusions about significant differences the same?

# Part 2: Variety

In a variety trial, a total of 7 varieties (Var) of wheat are considered. For each variety, n=3 reps are randomly assigned to positions in a single field. The response variable is the Yield at the end of the trial. The data is available from Canvas as “Varieties.csv”. _Note: You may need to detach the emmeans package and then reload it for this group of questions._

## Question 9

Treat Variety as fixed and fit a one-way ANOVA model using `lm()`. Fit an appropriate model and construct the ANOVA table (using `Anova(, type = 3)`).

## Question 10

In the ANOVA table, you should find a test corresponding to Variety. Specifically, F = 18.26 and p-value <.0001. State the null hypothesis for this test.

## Question 11

Give the emmeans for each variety.

## Question 12

Treat Variety as random and fit a one-way random effects model using `lmer()`. Fit an appropriate model and include “Random Effects” table (giving the variance component estimates) in your assignment.

## Question 13

Use the `rand()` function to get a test corresponding to Variety. Specifically, Chi.sq = 19 and p-value < 0.0001. State the null hypothesis for this test.

## Question 14

Give the BLUPs for each variety. Remember to include the intercept in the BLUP estimation.

\newpage

# Appendix

```{r show_code, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
