---
title: "STAT 512 Homework 5"
author: "Kathleen Wendt"
date: "03/24/2020"
output: pdf_document
---

```{r packages, include = FALSE}
# load packages
library(tidyverse)
library(janitor)
# library(car)
# library(emmeans)
# library(broom)
# library(kableExtra)
# library(MuMIn)
```

```{r global_options, include = FALSE}
# set global options
knitr::opts_chunk$set(fig.width = 6, 
                      fig.height = 4, 
                      fig.path = "figs/",
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

# Part 1: Heart Data

For this problem use the data described in Ott and Longnecker Example 12.22 (p 664 in the 7th edition). The data are available from Canvas as “CKheart.csv”. Read the description of the data in the book. You can use the output in the book to check your own R calculations.

```{r heart_data}
# read heart data
heart_data <- readr::read_csv("data/CKheart.csv") %>% 
  janitor::clean_names()
```

## Question A: Logistic Regression

Use `glm()` to fit a logistic regression model that estimates the probability of a heart attack as a function of CK value. Include the Coefficients table in your assignment.

## Question B: Plot

Construct a plot of the data with the fitted logistic regression curve overlaid. Include the plot in your assignment.

## Question C: OR / CI

Give an estimate of the odds ratio corresponding to CK and an approximate 95% confidence interval.

## Question D: CK Odds

Give a one-sentence description of the odds of heart attack among those with a given level of CK, compared to the odds of a heart attack among those with a level of CK ten points higher. (4 pts)

## Question E: Psuedo R^2

Calculate McFadden’s pseudo R2 for the model.

## Question F: LD90

Give an estimate of the CK level at which doctors would be 90% sure that a subject has had a heart attack.

# Part 2: Birth Weight

An observational study was done to investigate risk factors associated with low infant birth weight. Data from 189 (singleton) pregnancies were collected at Baystate Medical Center, Springfield, MA during 1986. The response variable was low (1 if birth weight was less than 2.5 kg, 0 otherwise). The predictor variables included: age (mother’s age in years), mwt (mother’s weight in pounds prior to pregnancy), race (mother’s race, 1= white, 2=black, 3=other) and smoke (1=mother smoked during pregnancy, 0 otherwise). The data is available from Canvas as “birthweight.csv”.

Important note: Be sure to define race and smoke as factors!

## Question A: Race

To examine the relationship between low vs race: calculate the proportion of births resulting in low birthweight for each race category and present the p-value from a chi-square test. (4 pts)

## Question B: Smoking

To examine the relationship between low vs smoke: calculate the proportion of births resulting in low birthweight for each smoke category and present the p-value from a chi-square test. (4 pts)

## Question C: Logistic Regression

Run a logistic regression with smoke as the only predictor variable. Calculate the emmeans using type = “response” for each smoke group (copy/paste the results to your assignment). Note: these should match your simple proportions from part B. (4 pts)

## Question D: AIC Model Selection

Now consider all 4 predictors (age, mwt, race, smoke). Using best subsets selection with AIC criteria, which variables are included in the final model? Include the Coefficients table and Type3 Anova table in your assignment. (4 pts)

_NOTE: Use the selected model from the previous question for all further questions!_

## Question E: OR / CI

Based on the model selected above, give the estimated odds ratio and
corresponding 95% CI for Smokers vs Non-Smokers (smoke 1 vs 0).

## Question: F: EM Means

Calculate the emmeans using type = “response” for each smoke group
(copy/paste the results to your assignment). Note that these values are different from what you found in part C because of the additional variables included in the model.

## Question G: Tukey

Run Tukey adjusted pairwise comparisons for race. Discuss your findings. (4
pts)

## Question H: Hoslem Test

Give the p-value corresponding to the Hosmer-Lemeshow test. Use `hoslem.test()` from the `ResourceSelection` package with g = 10 groups. Based on this test, is there evidence of lack of fit?

\newpage

# Appendix

```{r show_code, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
