---
title: "STAT 512 Homework 3"
author: "Kathleen Wendt"
date: "02/18/2020"
output: pdf_document
---

```{r packages, include = FALSE}
# load packages
library(tidyverse)
library(janitor)
library(broom)
library(car)
```

```{r global_options, include = FALSE}
# set global options
knitr::opts_chunk$set(fig.width = 6, 
                      fig.height = 4, 
                      fig.path = "figs/",
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

# Part A: Grain 

Questions 1 through 6 (Grain Yield): Data relating grain Yield (Y) to the number of Days (X) after flowering that harvesting took place was examined in “Determination of Biological Maturity and Effect of Harvesting and Drying Conditions on Milling Quality of Paddy” (J of Ag Engr. Research (1975):353-361.) The data is available from Canvas as “Grain.csv”.

Notes:

- For consistency, please use the `I()` or `poly( , raw = TRUE)` functions for fitting the quadratic and cubic models.

- For questions 2-4, you do NOT need to include the diagnostic plots in your assignment. Just discuss your findings.

```{r grain_data}
# read grain data
grain_data <- readr::read_csv("data/Grain.csv") %>% janitor::clean_names()
```

## Question 1: Scatterplot

Create a scatterplot of Yield vs Days. Include this plot in your assignment.

```{r grain_plot}
# 1. pairwise plot (base) for grain data
plot(grain_data)
```

## Question 2: Simple linear regression

Fit a linear regression model of Yield on Days. Include the parameter estimate information (“Coefficients” table) in your assignment. Examine a plot of the residuals versus predicted values. What does the residual plot suggest? (4 pts)

```{r grain_lm}
# 2. grain simple lin reg 
grain_lm <- lm(yield ~ days, data = grain_data)
summary(grain_lm)
```

```{r grain_lm_plot, include = FALSE}
# 2. res vs. fitted plot of grain simple lin reg model 
plot(grain_lm, which = 1)
```

The plot of the residuals vs. predicted values for the simple linear regression of yield (Y) on days (X) revealed curvature, indicating a violation of the assumption of linearity.

## Question 3: Quadratic regression model

Fit a quadratic regression model (including both linear and quadratic terms). Include the parameter estimate information (“Coefficients” table) in your assignment. Examine a plot of the residuals versus predicted values and comment. (4 pts)

```{r grain_quadlm}
# 3. grain quad reg model
grain_quadlm <- lm(yield ~ days + I(days^2), data = grain_data)
summary(grain_quadlm)
```

```{r grain_quadlm_plot, include = FALSE}
# 3. res vs. fitted plot of grain quadratic model 
plot(grain_quadlm, which = 1)
```

The plot of the residuals vs. predicted values for the quadratic regression of yield (Y) on days (X) with both linear (X) and quadratic terms of days ($X^2$) showed an improvement in scatter, although non-constant variance might be a slight issue.

## Question 4: Cubic regression model

Fit a cubic regression model (including linear, quadratic, and cubic terms). Include the parameter estimate information (“Coefficients” table) in your assignment. Again examine a plot of the residuals versus predicted values and comment. (4 pts)

```{r grain_cubiclm}
# 4. grain cubic reg model
grain_cubiclm <- lm(yield ~ days + I(days^2) + I(days^3), data = grain_data)
summary(grain_cubiclm)
```

```{r grain_cubiclm_plot, include = FALSE}
# 4. res vs. fitted plot of grain quadratic model 
plot(grain_cubiclm, which = 1)
```

The plot of the residuals vs. predicted values for the cubic regression of yield (Y) on days (X) with linear, quadratic, and cubic terms for days (X) is similar to the corresponding plot for the quadratic regression model.

## Question 5: Hypothesis test for cubic model

In the cubic model (#4), test the hypothesis that the linear, quadratic and cubic regression coefficients are all simultaneously zero. Give the F-statistic and p-value and make a conclusion about the test.

```{r grain_nsht}
# 5. test if linear, quadratic, and cubic reg coeffs are 0
# 5. create matrix pattern
grain_matrix_q5 <- matrix(c(0, 1, 0, 0,
                            0, 0, 1, 0,
                            0, 0, 0, 1),
                          nrow = 3,
                          ncol = 4,
                          byrow = TRUE)
# 5. complete hypothesis test for coeffs
grain_betas <- broom::tidy(car::lht(grain_cubiclm,
                                    grain_matrix_q5,
                                    rhs = c(0, 0, 0)))
```

Based on the linear hypothesis test, we reject the null hypothesis. At least one of the partial regression coefficients (linear, quadratic, cubic of `days`) in the cubic model for grain yield is non-zero, _F_ = `r round(grain_betas$statistic[2], 3)`, _p_ = `r round(grain_betas$p.value[2], 5)` < $\alpha$ = 0.05.

## Question 6: Model selection

Which model would you choose: linear, quadratic or cubic? Justify your choice. Hint: Think about the simplest model that satisfies assumptions.

I would select the _quadratic_ model because it is the simplest model that satisfies assumptions reasonably well and seems to fit the data well ($R^2$ = 0.79).

\newpage

# Part B: Drugs

Data was collected to compare 3 drug treatments for leprosy. Three variables are included in the data set: 

- `Drug`: drug treatment (A, D or F)

- `PreTreatment`: a pretreatment score of leprosy bacilli

- `PostTreatment`: a posttreatment score of leprosy bacilli

Ten patients were randomly assigned to each Drug treatment (for a total of n = 30 subjects). The goal of the study is to compare mean PostTreatment values for the 3 drugs. Hence, PostTreatment is the response variable and Drug is the predictor variable. But the researchers would like to consider including the PreTreatment value as a covariate. The data is available from Canvas as “DrugTest.csv”.

```{r drug_data}
# read drug data
drug_data <- readr::read_csv("data/DrugTest.csv") %>% 
  janitor::clean_names() %>% 
  dplyr::mutate(drug = as.factor(drug)) 
```

## Question 7: Plots

Construct side-by-side boxplots of (1) PreTreatment vs Drug and (2) PostTreatment vs Drug. Also construct (3) a scatterplot of PostTreatment vs PreTreatment for all Drugs on the same plot. Overlay a fitted regression line for each Drug. Include these plots in your assignment. (4 pts)

```{r drug_prebox}
# 7. side-by-side boxplot of pre by drug
drug_data %>% 
  dplyr::group_by(drug) %>% 
  ggplot2::ggplot(aes(x = drug, y = pre_treatment, color = drug)) +
  geom_boxplot() +
  ggtitle("Pre-treatment levels by drug type") +
  theme_minimal()
```

```{r drug_postbox}
# 7. side-by-side boxplot of post by drug
drug_data %>% 
  dplyr::group_by(drug) %>% 
  ggplot2::ggplot(aes(x = drug, y = post_treatment, color = drug)) +
  geom_boxplot() +
  ggtitle("Post-treatment levels by drug type") +
  theme_minimal()
```

```{r drug_scatter}
# 7. scatter plot of pre vs. post by drug
drug_data %>% 
  dplyr::group_by(drug) %>% 
  ggplot2::ggplot(aes(x = pre_treatment, y = post_treatment, color = drug)) +
  geom_point() +
  geom_smooth(formula = "y ~ x", method = "lm", fill = NA) +
  ggtitle("Patterns in pre- to post-treatment levels by drug type") +
  theme_minimal()
```

## Question 8: One-way ANOVA

Fit the one-way ANOVA model (using Drug as the only predictor). Include the ANOVA table and Tukey adjusted pairwise comparisons in your assignment. What can we conclude about differences between the Drugs? (4 pts)

```{r drug_anova}
# 8. drug one-way anova
```

## Question 9

Now fit the ANCOVA model with NO Interaction (using Drug and PreTreatment as the predictors). Include the ANOVA table and Tukey adjusted pairwise comparisons in your assignment. What can we conclude about differences between the Drugs? (4 pts)

## Question 10

Comparing your conclusions from #8 (one-way ANOVA) vs #9 (ANCOVA) you should
have found different conclusions regarding significant differences between the Drug treatments. Give a brief explanation of why the conclusions change when we include PreTreatment as a covariate. Hint: Consider the boxplots from #7.

## Question 11

An alternative approach to the ANCOVA above is to calculate the difference (Diff = PostTreatment – PreTreatment) and use this Diff as the response in a one-way ANOVA model. Do this and include the ANOVA table and Tukey adjusted pairwise comparisons in your assignment. What can we conclude about differences between the Drugs? (4 pts)

\newpage

# Appendix

```{r show_code, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
