---
title: "STAT 512 Homework 8"
author: "Kathleen Wendt"
date: "04/14/2020"
output: pdf_document
---

```{r packages, include = FALSE}
# load packages
library(tidyverse)
library(janitor)
library(kableExtra)
library(car)
library(broom)
library(emmeans)
```

```{r global_options, include = FALSE}
# set global options
knitr::opts_chunk$set(fig.width = 6, 
                      fig.height = 4, 
                      fig.path = "figs/",
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

# Biomass Study

A greenhouse study was done to examine the effect of 3 herbicides (A,B, or C) and 2 water regimes (1 = Low or 2 = High) for 2 plant types (Grass or Forb). The response variable is biomass. There are 3 reps per treatment combination for a total of 36 observations. Each observation was a potted plant. The 36 pots were randomly assigned without restriction to locations in the greenhouse. The data is available from Canvas as “Biomass.csv”.

```{r bio_data}
# read and prepare biomass data
bio_data <- readr::read_csv("data/Biomass.csv") %>%
  janitor::clean_names() %>% 
  dplyr::mutate(type = as.factor(type), 
                herb = as.factor(herb),
                water = as.factor(water))
```

## Question 1: Plots

Create two-way interaction plots separately for each level of Type. (4 pts) This can be done by first calculating treatment means (using aggregate or dplyr).

```{r 1_sumstats}
# 1. build summary stats by type
bio_table <- bio_data %>%
  dplyr::group_by(type, herb, water) %>%
  dplyr::summarize(
    n = n(),
    mean_biomass = mean(biomass),
    sd_biomass = sd(biomass)
    )
```

``` {r 1_plots}
# 1. build two-way interaction plots
ggplot2::qplot(x = herb,
               y = mean_biomass,
               color = water,
               group = water,
               data = bio_table) +
  geom_line() +
  facet_grid(. ~ type)
```

## Question 2: Three-way ANOVA model

Fit the three-way model with all interactions. Include the Type 3 ANOVA table in your assignment. You should find evidence of a 3 way interaction.

```{r 2_bio_3_anova}
# 2. adjust contrast default
options(contrasts = c("contr.sum", "contr.poly"))
# 2. fit three-way anova model
bio_anova_3 <- lm(biomass ~ type*herb*water,
                 data = bio_data)
# 2. call tidy anova type 3 table
kableExtra::kable(broom::tidy(car::Anova(bio_anova_3, type = 3)))
```

There is evidence supporting a three-way interaction between plant type, herbicide, and water pattern.

## Question 3: Assumptions

Consider the diagnostics plots and discuss whether model assumptions are satisfied.

```{r 3_diagnostics, include = FALSE}
# 3. diagnostic plots for three-way anova model
plot(bio_anova_3)
```

Residuals vs Fitted: Residuals in megaphone pattern and not evenly distributed. Assumptions of linearity and constant variance likely not supported.

Normal Q-Q: Residuals do not tightly follow line. Heavy tails. Assumption of normality likely not supported, although this may be due in part to the small sample size.

## Question 4: Pairwise comparisons

Use `emmeans` to calculate pairwise comparisons of Water (1 vs 2) for each level of Herb and Type. You can use code something like this:
`emmeans(Model1, pairwise ~ Water|Herb*Type)`.

```{r 4_pairs_water}
# 4. pairwise comparisons of water for each herb and type level/combo
bio_water_pairs <- emmeans::emmeans(bio_anova_3, pairwise ~ water|herb*type)
# 4. kable emmeans and contrasts
kableExtra::kable(bio_water_pairs$contrasts)
```

## Question 5: Forb two-way model

Now fit a two-way model (including interaction) for Forb only. Include the Type 3 ANOVA table in your assignment.

```{r 5_forb_anova}
# 5. subset data to forb type only
forb_data <- bio_data %>% dplyr::filter(type == "Forb")
# 5. adjust contrast default
options(contrasts = c("contr.sum", "contr.poly"))
# 5. fit two-way anova model for forb data
forb_anova <- lm(biomass ~ herb*water,
                 data = forb_data)
# 5. call tidy anova type 3 table
kableExtra::kable(broom::tidy(car::Anova(forb_anova, type = 3)))
```

No evidence for interaction between herbicide and water pattern for forb.

## Question 6: Assumptions

Consider the diagnostics plots and discuss whether model assumptions are satisfied.

```{r 6_diagnostics, include = FALSE}
# 6. diagnostic plots for two-way anova model of type forb
plot(forb_anova)
```

Residuals vs Fitted: Residuals evenly distributed across level but still cluster. Assumptions of linearity and constant variance seem reasonable, but there systematic differences in variance across levels.

Normal Q-Q: Residuals close to line. Assumption of normality is supported.

## Question 7: Pairwise comparisons

Use `emmeans` to calculate pairwise comparisons of Water (1 vs 2) for each level of Herb.

```{r 7_pairs_water}
# 7. pairwise comparisons of water for each herb level
forb_water_pairs <- emmeans::emmeans(forb_anova, pairwise ~ water|herb)
# 7. kable emmeans and contrasts
kableExtra::kable(forb_water_pairs$contrasts)
```

## Question 8: Forb LSD

Calculate the LSD value (ME for difference between means) corresponding to the comparisons from the previous question. Show your work for full credit. _Hint: See ExpDesign2 slide 81. Recall that you can check your calculation using the SE value given in the output from the previous question._

LSD = 0.40

See code appendix for hand calculation.

```{r 8_forb_lsd, include = FALSE}
# 8. calculate t critical value in R
qt(0.975, 12)
# 8. long-hand
2.178*(sqrt((2*(0.6268/12)/3)))
# 8. short-cut with se from contrasts output
2.178*0.187
```

## Question 9: Pairwise comparisons

Use `emmeans` to calculate the comparison of Water (1 vs 2) averaging over the levels of Herb.

```{r 9_pairs_water}
# 9. pairwise comparisons of water averaging over herb 
forb_water_pairs_avg <- emmeans::emmeans(forb_anova, pairwise ~ water)
# 9. kable emmeans and contrasts
kableExtra::kable(forb_water_pairs_avg$contrasts)
```

## Question 10: Forb (avg) LSD

Calculate the LSD value (ME for difference between means) corresponding to the comparisons from the previous question. Show your work for full credit. 

LSD = 0.23

See code appendix for hand calculation.

```{r 10_forb_avg_lsd, include = FALSE}
# 10. calculate t critical value in R
qt(0.975, 12)
# 10. long-hand 
2.178*(sqrt((2*(0.6268/12)/9)))
# 10. short-cut with se from contrasts output
2.178*0.108
```

## Question 11: Power

Comparing your LSD values from questions #8 (interaction comparison) and #10 (main effect comparison), which comparison has higher power? Briefly discuss.

As sample size increases and margin of error decreases, power increases; therefore, the main effect comparison (#10) has higher power. By averaging across herbicide, there were more degrees of freedom in the second model.

## Question 12: Grass two-way model 

Now fit a two-way model (including interaction) for Grass only. Include the Type 3 ANOVA table in your assignment.

```{r 12_grass_anova}
# 12. subset data to grass type only
grass_data <- bio_data %>% dplyr::filter(type == "Grass")
# 12. adjust contrast default
options(contrasts = c("contr.sum", "contr.poly"))
# 12. fit two-way anova model for grass data
grass_anova <- lm(biomass ~ herb*water,
                 data = grass_data)
# 12. call tidy anova type 3 table
kableExtra::kable(broom::tidy(car::Anova(grass_anova, type = 3)))
```

Evidence of interaction between herbicide and water pattern for grass.

## Question 13: Assumptions

Consider the diagnostics plots and discuss whether model assumptions are satisfied.

```{r 13_diagnostics, include = FALSE}
# 13. diagnostic plots for two-way anova model of type grass
plot(grass_anova)
```

Residuals vs Fitted: Residuals evenly distributed across residual values but not fitted values. Assumptions of linearity and constant variance seem reasonable, but there there may be evidence of a systemic difference between the "two sides" of the data (i.e., interaction effect).

Normal Q-Q: Residuals close to line but tails look heavy. Assumption of normality is okay, given the small sample size.

## Question 14: Pairwise comparisons

Use `emmeans` to calculate pairwise comparisons of Water (1 vs 2) for each level of Herb.

```{r 14_pairs_water}
# 14. pairwise comparisons of water for each herb level
grass_water_pairs <- emmeans::emmeans(grass_anova, pairwise ~ water|herb)
# 14. kable emmeans and contrasts
kableExtra::kable(grass_water_pairs$contrasts)
```

## Question 15: Reflection

Would it be appropriate to calculate the comparison of Water (1 vs 2) averaging over the levels of Herb? Briefly discuss.

No, it would not be appropriate to average over herbicide because there is evidence of an interaction between water pattern and herbicide among plant type `grass`. 

## Question 16: Model comparison

Now we compare the three-way model to the separate two-way models. The comparisons from #7 and #14 were also considered in #4. Considering the corresponding results from the combined analysis (#4) versus the split analysis (#7, #14), briefly discuss what is the same and what is different. Specifically, consider the estimated difference, corresponding SE and df. (4 pts)

The combined model (#4) had 24 degrees of freedom, whereas the split-type models (#7 and #14) each had 12 degrees of freedom. Split model 1 for `forb` (#7) had the smallest standard errors; split model 2 for `grass` (#14) had the largest standard errors. The estimated marginal means and estimated differences across the models (i.e., combined vs. split 1; combined vs. split 2) were the same.

## Question 17: Reflection

Give (at least) one reason why we might prefer to split the analysis by Type (running separate 2way ANOVAs for Grass and Forb). Your answer should be based on the output.

Forb and Grass had very different patterns of mean biomass by water pattern and herbicide. Splitting the data would simplify the analysis and specify comparisons of interest consistent with the research questions (e.g., maybe huge differences by plant type are already known). Also, splitting the data might help with the concern of unequal variance.

## Question 18: Reflection

Give (at least) one weakness of splitting the analysis “by type” as compared to the full 3-way ANOVA model.

Splitting the data reduces dfResid, thereby reducing power. In this case, the three-way model had 24 dfResid, but each two-way model only had 12 dfResid.

\newpage

# Appendix

```{r show_code, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
